---
title: "00.1: Organism EDA"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries
```{r}

library(tidyverse)
library(vegan)
library(ggplot2)
library(reader)

```

# Load data
```{r}

# figure out where you are. you'll need to know so you can adjust the path in the data import lines below.
getwd()

# add data
data23 <- read.csv("../data/yi2023_completeNK.csv")
data24 <- read.csv("../data/yi2024total.csv")
print(data23)
print(data24)

```

# Data cleaning
## fix capitalization, aggregate counts across zone, pivot to usable format
```{r}

# fix column names
colnames(data23) <- tolower(colnames(data23))
colnames(data24) <- tolower(colnames(data24))

# aggregate counts + pivot
clean_data <- function(data, year) {
  # Consolidate counts by summing within each zone
  data_cleaned <- data %>%
    group_by(zone) %>%
    summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>%
    pivot_longer(-zone, names_to = "organism", values_to = "count") %>%
    pivot_wider(names_from = "zone", values_from = "count", values_fill = 0)
  
  # Add year column
  data_cleaned <- data_cleaned %>%
    mutate(year = year)
  
  return(data_cleaned)
}

# create cleaned df and write out to tables
cleaned23 <- clean_data(data23, 2023)
cleaned24 <- clean_data(data24, 2024)

#write.csv(cleaned23, "/Users/cmantegna/Documents/GitHub/sicb_nk/output/data_adjusted/cleaned23.csv", row.names = FALSE)
#write.csv(cleaned24, "/Users/cmantegna/Documents/GitHub/sicb_nk/output/data_adjusted/cleaned24.csv", row.names = FALSE)

```

## adjust dfs for visualization
## NOTE: cleaned up the naming of some catagories mannually
```{r}

# rename your cleaned files back to data23 and data24 so you aren't constantly re-entering new file/ object names for each step, that's a surefire way to get confused.
data23<- cleaned23
data24<- cleaned24

# combine them for a single file for comparisons
presence_absence <- function(data23, data24) {
  data_combined <- data23 %>%
    full_join(data24, by = "organism", suffix = c("_2023", "_2024")) %>%
    mutate(across(ends_with("2023"), ~ ifelse(. > 0, 1, 0), .names = "binary_{col}"),
           across(ends_with("2024"), ~ ifelse(. > 0, 1, 0), .names = "binary_{col}"))
  return(data_combined)
}

comparison <- presence_absence(data23, data24)

# check for NA values since we went from 68 obs. in 2023 and 67 obs. in 2024 to a combined 73 obs, in the comparison. 
# backfill NAs

# replace NAs
comparison <- comparison %>%
  mutate(
    across(c(L_2023, L_2024, M_2023, M_2024, U_2023, U_2024), ~ replace_na(., 9999)),
    year_2024 = replace_na(year_2024, 2024),
    year_2023 = replace_na(year_2023, 2023)
  )

# write out df 
write.csv(comparison, "../data_adjusted/comparison.csv", row.names = FALSE)

```

# EDA
## what we're going for
```{r}

# Number of taxa present for each year
# Number of identified species for each year
# Comparison of taxa and species between 2023 & 2024

```


## visualize
## NOTE: 2024 upper-level species count feels abnormally high. must go back and double check dataframes to see what's driving that.
## NOTE: organism distribution plots will need to be done in groups so that the barnacles and limpets/ other snails don't blow it out.

```{r}

# total count by tidal zone
plot_zone_distribution <- function(comparison, year) {
  data_long <- comparison %>%
    pivot_longer(-c(organism, year), names_to = "zone", values_to = "count")
  
  ggplot(data_long, aes(x = zone, y = count, fill = zone)) +
    geom_bar(stat = "identity") +
    labs(title = paste("Total Organism Counts by Zone -", year),
         x = "Zone", y = "Total Counts") +
    theme_minimal()
}

plot_zone_distribution(data23, 2023)
plot_zone_distribution(data24, 2024)

library(ggplot2)
library(dplyr)

# Function to plot organism distributions based on thresholds
plot_organism_distribution <- function(data, year, thresholds) {
  data_long <- data %>%
    rowwise() %>%
    mutate(total_count = sum(c_across(-c(organism, year)), na.rm = TRUE)) %>%
    ungroup() %>%
    filter(total_count > 0)  # Exclude organisms with a total count of 0
  
  # Split data into categories based on thresholds
  data_1 <- data_long %>% filter(total_count > thresholds[[1]][1], total_count <= thresholds[[1]][2])
  data_2 <- data_long %>% filter(total_count > thresholds[[2]][1], total_count <= thresholds[[2]][2])
  data_3 <- data_long %>% filter(total_count > thresholds[[3]][1])  # No upper limit for the last category
  
  # Plot organisms in each category
  plot_1 <- ggplot(data_1, aes(x = reorder(organism, total_count), y = total_count)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    coord_flip() +
    labs(title = paste("Organisms with Counts in Range", paste(thresholds[[1]], collapse = "-"), "-", year),
         x = "Organism", y = "Total Counts") +
    theme_minimal()
  
  plot_2 <- ggplot(data_2, aes(x = reorder(organism, total_count), y = total_count)) +
    geom_bar(stat = "identity", fill = "darkorange") +
    coord_flip() +
    labs(title = paste("Organisms with Counts in Range", paste(thresholds[[2]], collapse = "-"), "-", year),
         x = "Organism", y = "Total Counts") +
    theme_minimal()
  
  plot_3 <- ggplot(data_3, aes(x = reorder(organism, total_count), y = total_count)) +
    geom_bar(stat = "identity", fill = "green") +
    coord_flip() +
    labs(title = paste("Organisms with Counts >", thresholds[[3]][1], "-", year),
         x = "Organism", y = "Total Counts") +
    theme_minimal()
  
  return(list(category_1 = plot_1, category_2 = plot_2, category_3 = plot_3))
}

# Define thresholds for 2023 and 2024
thresholds_2023 <- list(c(0, 30), c(31, 100), c(101, Inf))
thresholds_2024 <- list(c(0, 60), c(61, 500), c(501, Inf))

# Generate plots for 2023
plots_2023 <- plot_organism_distribution(data23, 2023, thresholds_2023)

# Generate plots for 2024
plots_2024 <- plot_organism_distribution(data24, 2024, thresholds_2024)

# Display the plots for 2023
print("Plots for 2023:")
print(plots_2023$category_1)  # Organisms with counts 1-50
print(plots_2023$category_2)  # Organisms with counts 51-100
print(plots_2023$category_3)  # Organisms with counts 101+

# Display the plots for 2024
print("Plots for 2024:")
print(plots_2024$category_1)  # Organisms with counts 1-100
print(plots_2024$category_2)  # Organisms with counts 101-500
print(plots_2024$category_3)  # Organisms with counts 501+

# 2023 - 2024 comparison
comparison_visualization <- function(comparison) {
  # Calculate the difference between 2024 and 2023 for each organism
  comparison <- comparison %>%
    mutate(difference = binary_count_2024 - binary_count_2023) %>%
    filter(!is.na(difference)) # Remove any rows where difference is NA
  
  # Create the visualization
  ggplot(comparison, aes(x = reorder(organism, difference), y = difference)) +
    geom_bar(stat = "identity", fill = "coral") +
    coord_flip() +
    labs(title = "Difference in Organism Counts Between Years",
         x = "Organism", y = "Count Difference (2024 - 2023)") +
    theme_minimal()
}

# Call the function with the corrected dataset
comparison_visualization(comparison)


```

# next steps
```{r}

# alpha and beta diversity
# species richness
# t-test between years (total), and zonation (L and M), species level
# for stats: PERMANOVA
# box plots/ stacked bars (high level taxa distribution)
```

```{r}
print(data23)
print(data24)

plot_zone_distribution_limpet <- function(comparison, year) {
  data_long <- comparison %>%
    pivot_longer(-c(organism, year), names_to = "zone", values_to = "count") %>%
    filter(organism == "limpet") # Filter for limpets only
  
  ggplot(data_long, aes(x = zone, y = count, fill = zone)) +
    geom_bar(stat = "identity") +
    labs(title = paste("Limpet Counts by Zone -", year),
         x = "Zone", y = "Total Counts") +
    theme_minimal()
}

plot_zone_distribution_limpet(data23, 2023)
plot_zone_distribution_limpet(data24, 2024)

plot_zone_distribution_barnacles <- function(comparison, year) {
  data_long <- comparison %>%
    pivot_longer(-c(organism, year), names_to = "zone", values_to = "count") %>%
    filter(organism == "barnacles") # Filter for limpets only
  
  ggplot(data_long, aes(x = zone, y = count, fill = zone)) +
    geom_bar(stat = "identity") +
    labs(title = paste("Barnacle Counts by Zone -", year),
         x = "Zone", y = "Total Counts") +
    theme_minimal()
}

plot_zone_distribution_barnacles(data23, 2023)
plot_zone_distribution_barnacles(data24, 2024)


plot_zone_distribution_anemone <- function(comparison, year) {
  data_long <- comparison %>%
    pivot_longer(-c(organism, year), names_to = "zone", values_to = "count") %>%
    filter(organism == "anemone") # Filter for anemone's only
  
  ggplot(data_long, aes(x = zone, y = count, fill = zone)) +
    geom_bar(stat = "identity") +
    labs(title = paste("anemone Counts by Zone -", year),
         x = "Zone", y = "Total Counts") +
    theme_minimal()
}

plot_zone_distribution_anemone(data23, 2023)
plot_zone_distribution_anemone(data24, 2024)
```
```{r}
#species_richness
plot_species_count_by_zone <- function(data, year) {
  # Calculate the number of species in each zone
  data_species_count <- data %>%
    pivot_longer(-c(organism, year), names_to = "zone", values_to = "count") %>%
    group_by(zone) %>%
    summarise(species_count = sum(count > 0, na.rm = TRUE)) # Count species present in each zone
  
  # Create the bar plot
  ggplot(data_species_count, aes(x = zone, y = species_count, fill = zone)) +
    geom_bar(stat = "identity") +
    labs(title = paste("Number of Species by Zone -", year),
         x = "Zone", y = "Number of Species") +
    theme_minimal()
}

# Plot for 2023 data
plot_species_count_by_zone(data23, 2023)

# Plot for 2024 data
plot_species_count_by_zone(data24, 2024)
```

```{r}
# Load required libraries
library(tidyverse)



# Add year column to differentiate datasets
data23 <- data23 %>% mutate(year = 2023)
data24 <- data24 %>% mutate(year = 2024)

# Combine datasets for analysis
combined_data <- bind_rows(data23, data24)

# T-Test Between Years (Total) for Each Species
perform_t_test_years <- function(data) {
  t_test_results <- data %>%
    pivot_longer(-c(organism, year), names_to = "zone", values_to = "count") %>%
    group_by(organism) %>%
    summarise(
      p_value = tryCatch(
        t.test(count[year == 2023], count[year == 2024], var.equal = TRUE)$p.value,
        error = function(e) NA
      )
    ) %>%
    arrange(p_value)
  return(t_test_results)
}

# T-Test Between Zones (L, M, U) for Each Species
perform_t_test_zones <- function(data, year) {
  t_test_results <- data %>%
    filter(year == year) %>%
    pivot_longer(-c(organism, year), names_to = "zone", values_to = "count") %>%
    group_by(organism) %>%
    summarise(
      p_value = tryCatch(
        t.test(count[zone == "L"], count[zone == "M"], var.equal = TRUE)$p.value,
        error = function(e) NA
      )
    ) %>%
    arrange(p_value)
  return(t_test_results)
}

# Run T-Test for Years
t_test_years <- perform_t_test_years(combined_data)
print("T-Test Results Between Years")
print(t_test_years)

```


```{r}
library(tidyverse)
library(vegan)

# Load the datasets
data23 <- read.csv("../data/yi2023_completeNK.csv")
data24 <- read.csv("../data/yi2024total.csv")

# Ensure column names are lowercase for consistency
colnames(data23) <- tolower(colnames(data23))
colnames(data24) <- tolower(colnames(data24))

# Define a cleaning function for the data
clean_data <- function(data, year) {
  data_cleaned <- data %>%
    group_by(zone) %>%
    summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>%
    pivot_longer(-zone, names_to = "organism", values_to = "count") %>%
    pivot_wider(names_from = "zone", values_from = "count", values_fill = 0) %>%
    mutate(year = year) # Add year column
  
  return(data_cleaned)
}

# Apply the cleaning function
cleaned23 <- clean_data(data23, 2023)
cleaned24 <- clean_data(data24, 2024)

# Combine datasets
comparison <- bind_rows(cleaned23, cleaned24)

# Check column names for debugging
print("Column names in 'comparison':")
print(colnames(comparison))

# Filter for limpets only
limpet_data <- comparison %>% filter(organism == "limpet")

# Debugging: Check the structure of 'limpet_data'
print("Structure of 'limpet_data':")
str(limpet_data)

# Transform the data to a long format for zones
# Use debugging to identify correct column names
limpet_long <- comparison %>%
    pivot_longer(-c(organism, year), names_to = "zone", values_to = "count") %>%
    filter(organism == "limpet") # Filter for limpets only

# Ensure no missing values
limpet_long <- limpet_long %>% filter(complete.cases(.))

# Prepare the species matrix
species_matrix <- limpet_long %>%
  pivot_wider(names_from = zone, values_from = count, values_fill = 0) %>%
  select(-year) %>%
  as.matrix()

# Prepare metadata
metadata <- limpet_long %>%
  select(year, zone) %>% 
  filter(complete.cases(.))

# Compute Bray-Curtis distance matrix
distance_matrix <- vegdist(species_matrix, method = "bray")

# Perform PERMANOVA test
permanova_result <- adonis2(
  distance_matrix ~ year * zone,
  data = metadata
)
print(colnames(limpet_data))
# Display results
print(permanova_result)
```





