---
title: "00.1: Algal EDA"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load libraries
```{r}

library(dplyr)
library(tidyverse)
library(vegan)
library(ggplot2)
library(reader)

```

# Load Data
```{r}

# figure out where you are on your computer
getwd()

# Note you'll need to change the file path to match where it is on your computer
data<- read.csv("/Users/cmantegna/Documents/GitHub/sicb_km/data/algae.csv")
#data2<- read.csv("/Users/cmantegna/Documents/GitHub/sicb_km/output/data_adjusted/pivoted_data.csv")

```

# Clean it up
## convert presence/ absence to numerical values we can use
```{r}

# replace 'P' with 1 and 'A' with 0
data[, 3:ncol(data)] <- lapply(data[, 3:ncol(data)], function(x) ifelse(x == "P", 1, 0))

head(data)

```

## check for empties and ensure section is a character/ factor
```{r}

# fix column names
colnames(data) <- tolower(colnames(data))

# make sections characters & check they really are characters
data$section <- as.character(data$section)
str(data)

# check for missing values
colSums(is.na(data))

```

## transform the df
```{r}

library(dplyr)

data <- data %>%
  mutate(
    section = trimws(section),
    zonation = trimws(zonation)
  )

unique_combinations <- data %>%
  distinct(section, zonation)

print(unique_combinations)

# summarize by section & zone
library(dplyr)

zone_section_summary <- data %>%
  group_by(section, zonation) %>%
  summarise(across(where(is.numeric), sum))

# check it out
print(zone_section_summary)

```

# Species Richness
```{r}

# add a column for species richness (sum of all algae presence per row)
#data$richness <- rowSums(data[, 3:ncol(data)])

# summarize richness by section
richness_summary <- data %>%
  group_by(section) %>%
  summarise(averageRichness = mean(richness))

# Create a bar plot for species richness
richness <- ggplot(richness_summary, aes(x = factor(section), y = averageRichness, fill = factor(section))) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = custom_colors) +  # Apply the custom color palette
  labs(
    title = "Species Richness by Section",
    x = "Section",
    y = "Average Species Richness",
    fill = "Section"  # Legend title
  ) +
  theme_minimal(base_size = 14) +  # Clean theme with larger font size
  theme(
    panel.grid = element_blank(),             # Remove gridlines
    panel.border = element_blank(),           # No border around the plot
    axis.line = element_line(color = "black"),  # Retain X and Y axes
    legend.position = "right"                 # Position the legend to the right
  )

# Display the plot
print(richness)


#ggsave(filename= "/Users/cmantegna/Documents/GitHub/sicb_km/output/figures/algal_richness.png", plot=richness, width = 12, height = 10, dpi = 300)

```

# Algal presence by section - cleaning up naming and plot colors
```{r}

library(tidyr)
library(ggplot2)

# rename algae from the mix of scientific and common names to their common names
colnames(data)[3:16] <- c(
  "Rockweed", "Sea Lettuce","Turfweed", "Sargassum", "Turkish Washcloth", "Turkish Towel", "Iridescent Seaweed", "Sugar Kelp", "Crustose Coralline Algae", "Tar Spot Algae", "Other Red Algae", "Other Green Algae", "Other Brown Algae", "Unknown Algae")

# create a named vector mapping algae to their group
algae_group <- c(
  "Rockweed" = "Green Algae", "Sea Lettuce" = "Green Algae", "Other Green Algae" = "Green Algae",
  "Turfweed" = "Brown Algae", "Sargassum" = "Brown Algae", "Turkish Washcloth" = "Brown Algae", 
  "Sugar Kelp" = "Brown Algae", "Other Brown Algae" = "Brown Algae",
  "Turkish Towel" = "Red Algae", "Iridescent Seaweed" = "Red Algae", 
  "Crustose Coralline Algae" = "Red Algae", "Tar Spot Algae" = "Red Algae", "Other Red Algae" = "Red Algae",
  "Unknown Algae" = "Black"
)

# Define individual colors for each algae
algae_colors <- c(
  "Rockweed" = "#24693D",    # Dark green base color
  "Sea Lettuce" = "#3C8A55", 
  "Other Green Algae" = "#5AAD7A",  # Lighter shades of green
  "Turfweed" = "#663000", 
  "Sargassum" = "#855740", 
  "Turkish Washcloth" = "#996136", 
  "Sugar Kelp" = "#B88C74", 
  "Other Brown Algae" = "#D4B4A2", 
  "Turkish Towel" = "#990000", 
  "Iridescent Seaweed" = "#CC3333", 
  "Crustose Coralline Algae" = "#E06666", 
  "Tar Spot Algae" = "#F29999", 
  "Other Red Algae" = "#F7CCCC", 
  "Unknown Algae" = "#000000"  # Black
)

# Reshape data to long format
algae_long <- data %>%
  pivot_longer(cols = 3:16, names_to = "Algae", values_to = "Presence") %>%
  filter(Presence > 0)  # Exclude algae with no presence

# Reorder levels of the Algae factor
algae_long$Algae <- factor(
  algae_long$Algae,
  levels = c(
    "Unknown Algae",                           # Black
    "Turkish Towel", "Iridescent Seaweed", 
    "Crustose Coralline Algae", "Tar Spot Algae", "Other Red Algae",  # Red
    "Turfweed", "Sargassum", "Turkish Washcloth", 
    "Sugar Kelp", "Other Brown Algae",         # Brown
    "Rockweed", "Sea Lettuce", "Other Green Algae"  # Green
  )
)

# Create the stacked bar plot
algaeColor_ordered<- ggplot(algae_long, aes(x = factor(section), y = Presence, fill = Algae)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = algae_colors) +  # Apply individual algae colors
  labs(
    title = "Algal Presence by Section",
    x = "Section",
    y = "Presence",
    fill = "Algae Type"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    panel.grid = element_blank(),             # Remove gridlines
    axis.line = element_line(color = "black"), # Retain X and Y axes
    legend.position = "right",                # Position legend
    legend.title = element_text(size = 12),   # Adjust legend title size
    legend.text = element_text(size = 10)     # Adjust legend text size
  )

print(algaeColor_ordered)

#ggsave(filename= "/Users/cmantegna/Documents/GitHub/sicb_km/output/figures/algaeColor_ordered.png", plot=algaeColor_ordered, width = 12, height = 10, dpi = 300)

```


# Algal presence by section
```{r}

library(dplyr)
library(tidyr)
library(ggplot2)

# summarize algal presence by section
algal_summary <- data %>%
  group_by(section) %>%
  summarise(across(where(is.numeric), sum)) %>%
  pivot_longer(cols = -c(section,richness), names_to = "Algae", values_to = "Count")

# stacked bar plot
algae<- ggplot(algal_summary, aes(x = factor(section), y = Count, fill = Algae)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Algal Composition by Section",
    x = "Section",
    y = "Count of Presence",
    fill = "Algae Type"
  ) +
  theme_minimal()

#ggsave(filename= "/Users/cmantegna/Documents/GitHub/sicb_km/output/figures/algae_by_section.png", plot=algae, width = 12, height = 10, dpi = 300)

```

# Stats
## Normality check
```{r}

# data is not normally distributed so we have to run a PERMANOVA instead of an ANOVA
data %>%
  group_by(section) %>%
  summarise(p_value = shapiro.test(richness)$p.value)


```

## PERMANOVA
### troubleshooting code for the matrix creation: lines 170-196
```{r}

# make section a factor
data$section <- as.factor(data$section)

# pull out the algae columns
algal_data <- data[, 3:16]

# Check for missing values
sum(is.na(algal_data))

# Remove rows with missing values (if necessary)
algal_data <- na.omit(algal_data)

# Check data type of each column
str(algal_data)

# Check for missing values by column
colSums(is.na(algal_data))

# Check for missing values by row
rowSums(is.na(algal_data))

# Check if all values are numeric and non-negative
all(algal_data >= 0, na.rm = TRUE)

# Convert non-numeric columns to numeric (if needed)
#algal_data <- data[, 3:16]
#algal_data[] <- lapply(algal_data, as.numeric)

library(vegan)
# since this distance matrix is sensitive to rows with 0s all the way across, we are going to figure out if that exists before trying to rerun the distance matrix

zero_rows <- rowSums(algal_data) == 0

# Count the number of such rows
sum(zero_rows)

# If needed, inspect these rows
#data[zero_rows, ]

# Remove rows with all zeros
algal_data <- algal_data[!zero_rows, ]

# compute a distance matrix (e.g., Jaccard or Bray-Curtis)
distance_matrix <- vegdist(algal_data, method = "bray")

# Subset data to match non-zero rows in algal_data
non_zero_indices <- rowSums(data[, 3:16]) > 0  # Adjust column range as needed
data <- data[non_zero_indices, ]

# Check for negative values or missing entries in the distance matrix
#distance_matrix[distance_matrix < 0]  # Negative values
#sum(is.na(distance_matrix))           # Missing values


# Run PERMANOVA
permanova_result <- adonis2(distance_matrix ~ section, data = data, permutations = 999)

# View the results
print(permanova_result)

```

## posthoc for permanova is a pairwise test
### troubleshoot this later
```{r}

library(devtools)
#install.packages("devtools")
devtools::install_github("pmartinezarbizu/pairwiseAdonis")
library(pairwiseAdonis)


```

## NMDS to visualize which section(s) is driving the significant result
```{r}

library(vegan)

# Perform NMDS on the distance matrix
nmds_result <- metaMDS(vegdist(algal_data, method = "bray"), k = 2, trymax = 100)

# Plot NMDS with sections as groups
plot(nmds_result, type = "t")
ordihull(nmds_result, groups = data$section, draw = "polygon", col = "blue", label = TRUE)


```

## NMDS plot cleaned up a bit
```{r}

library(vegan)
library(ggplot2)
library(ggforce) 

# Extract NMDS scores
nmds_scores <- as.data.frame(scores(nmds_result))
nmds_scores$section <- data$section

# Calculate convex hulls for each group
hull_data <- nmds_scores %>%
  group_by(section) %>%
  slice(chull(NMDS1, NMDS2))  # Calculate convex hull points


# Define custom colors for each group of sections
custom_colors <- c(
  "1" = "#00AACC", "2" = "#33E4FF", "3" = "#66f0ff",  # Palette 1: Sections 1–3
  "4" = "#2ca02c", "5" = "#98df8a",                  # Palette 2: Sections 4–5
  "6" = "#663000", "7" = "#996136", "8" = "#cc9b7a"  # Palette 3: Sections 6–8
)


# Plot NMDS with axes and custom colors
algaeNMDS<- ggplot() +
  # Add convex hulls for each section
  geom_polygon(data = hull_data, aes(x = NMDS1, y = NMDS2, fill = section, group = section), alpha = 0.3) +
  # Add points for each sample
  geom_point(data = nmds_scores, aes(x = NMDS1, y = NMDS2, color = section), size = 3) +
  # Apply custom color palettes
  scale_fill_manual(values = custom_colors) +
  scale_color_manual(values = custom_colors) +
  # Aesthetic adjustments
  labs(
    title = "NMDS Plot of Algal Composition by Section",
    x = "NMDS1",
    y = "NMDS2",
    fill = "Section",
    color = "Section"
  ) +
  theme_minimal(base_size = 14) +  # Clean theme with larger font size
  theme(
    legend.position = "right",               # Keep legend on the right
    panel.grid = element_blank(),            # Remove all gridlines
    panel.border = element_rect(color = "black", fill = NA),  # Add border for axis
    axis.line = element_line(color = "black")  # Add x and y axes
  )


ggsave(filename= "/Users/cmantegna/Documents/GitHub/sicb_km/output/figures/algaeNMDS.png", plot=algaeNMDS, width = 12, height = 10, dpi = 300)

```

